Exact asymptotic expressions for local clustering in hyperbolic random graphs

### 2018-11-18 clustering_infinite_model.tex
final short pass before sending to Tobias and Nikolaos

Since $\Gamma^{+}(k-2\alpha, \xi_{\alpha,\nu})/\Gamma(k - 2\alpha) \sim k^{-(2\alpha+1)}$ as $k \to \infty$ \PvdH{Find reference.} we deduce that
\begin{equation}\label{eq:degree_distribution_P_asymptotics}
	\Prob{D_\Pcal = k} \sim 2\alpha (\xi_{\alpha,\nu})^{2\alpha} k^{-(2\alpha + 1)}
	\quad \text{as } k \to \infty.
\end{equation}

@@

Finally we need to deal with the terms in $\Delta_\Pcal(y(z))$ that involve the incomplete Beta-function. Let $a, c \in \R$, $\xi_{\alpha,\nu}, b >0$ positive real numbers. Using the integral definition of the incomplete B-function, the change of variables $s=1-t$ gives:
\begin{align*}
	\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} B^-(1-z;b,c) \dd z 
	&=\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_0^{1-z} t^{b-1} (1-t)^{c-1} \dd t \dd z \\
	&=\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_z^1 s^{c-1} (1-s)^{b-1} \dd s \dd z
\end{align*}
Then changing the order of integration and using the substitution $u = \xi_{\alpha,\nu}/z$  and recognizing the upper incomplete $\Gamma$-function yields
\begin{align*}
	&\hspace{-30pt}\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_z^1 s^{c-1} (1-s)^{b-1} \dd s \dd z\\
	&=\int_0^1 \int_0^s z^a e^{-\xi_{\alpha,\nu}/z} \dd z \, s^{c-1} (1-s)^{b-1} \dd s \\
	&=\int_0^1 \int_{\xi_{\alpha,\nu}/s}^\infty \xi_{\alpha,\nu}^{a+1} u^{-a-2} e^{-u} \dd u \, s^{c-1} (1-s)^{b-1} \dd s \\
	&= \xi_{\alpha,\nu}^{a+1} \int_0^1 \Gamma^+(-a-1,\xi_{\alpha,\nu}/s) s^{c-1} (1-s)^{b-1} \dd s.
		\numberthis \label{eq:integral_gamma_function}
\end{align*}
To compute this last integral we make use of the fact that the incomplete $\Gamma$-function has a representation in terms of Meijer's $G$-function (see Lemma~\ref{lem:gamma_meijer_G} in Appendix~\ref{sec:Meijer_G_functions})
\[
	\Gamma^+(-a-1,\xi_{\alpha,\nu}/s) = \MeijerG{2}{0}{1}{2}{1}{-a-1,0}{\frac{\xi_{\alpha,\nu}}{s}},
\] 
which holds for any $a\in \R$ and $s>0$ (that for a fixed second argument, the upper incomplete $\Gamma$-function is entire in the first argument, see \cite[pp. 899, 1032ff.]{gradshteyn2015table}). \PvdH{@Markus:What does entire mean?}  
We can now evaluate the integral in~\eqref{eq:integral_gamma_function} using several identities for Meijer's $G$-function. 
%formula 8.351.3, p. 
 %(see the wolfram functions website \url{http://functions.wolfram.com/06.06.26.0005.01}, it can also be found in the paper \url{https://arxiv.org/pdf/1507.05571}; , so we get the integral
First, inserting the expression for the incomplete Gamma-function into~\eqref{eq:integral_gamma_function} gives
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \int_0^1 s^{c-1} (1-s)^{b-1} \MeijerG{2}{0}{1}{2}{1}{-a-1,0}{\frac{\xi_{\alpha,\nu}}{s}} \dd s
\end{align*}
Next we apply the inversion identity for Meijer's $G$-function (see \cite[p. 209, 5.3.1.(9))]{erdelyi1953higher}) to get
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \int_0^1 s^{c-1} (1-s)^{b-1} \MeijerG{0}{2}{2}{1}{2+a,1}{0}{\frac{s}{\xi_{\alpha,\nu}}} \dd s
\end{align*}
This expression is actually the Euler transform of Meijer's $G$-function (see \cite[p. 214, 5.5.2.(5)]{erdelyi1953higher}) and (as the conditions $2+1<2(0+2)$ and $|\arg(\xi_{\alpha,\nu}^{-1})| < \frac{\pi}{2}$ (as $\xi_{\alpha,\nu}>0$) and $1-c-b<1-c$ (as $b>0$)  are satisfied) it equals
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \Gamma(b) \MeijerG{0}{3}{3}{2}{1-c,2+a,1}{0,1-c-b}{\xi_{\alpha,\nu}^{-1}}
\end{align*}
Using again the inversion identity for Meijer's $G$-function we now get
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \Gamma(b) \MeijerG{3}{0}{2}{3}{1,b+c}{c,-1-a,0}{\xi_{\alpha,\nu}}
\end{align*}
Finally, plugging in $a= 6\alpha-k-3$, $b=2\alpha$, $c=3-4\alpha$ we obtain
\begin{equation}\label{eq:integral_Delta_P_Beta}
	\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} B^-(1-z;b,c) \dd z =
	\xi_{\alpha,\nu}^{6a-k-2} \Gamma(2\alpha) \MeijerG{3}{0}{2}{3}{1,3-2\alpha}{3-4\alpha,-6\alpha+k+2,0}{\xi_{\alpha,\nu}}
\end{equation}
%(note that this agrees with the result in the old approach, after `bringing the power of $\xi$ into Meijer's $G$-function'.)

>>>>

Since $\Gamma^{+}(k-2\alpha, \xi_{\alpha,\nu})/\Gamma(k - 2\alpha) \sim k^{-(2\alpha+1)}$ as $k \to \infty$ \PvdH{@All: Does anyone have a good standard reference for this?} we deduce that
\begin{equation}\label{eq:degree_distribution_P_asymptotics}
	\Prob{D_\Pcal = k} \sim 2\alpha (\xi_{\alpha,\nu})^{2\alpha} k^{-(2\alpha + 1)}
	\quad \text{as } k \to \infty.
\end{equation}

@@

Finally we need to deal with the terms in $\Delta_\Pcal(y(z))$ that involve the incomplete Beta-function. Let $a, c \in \R$, $\xi_{\alpha,\nu}, b >0$ positive real numbers. Using the integral definition of the incomplete B-function, the change of variables $s=1-t$ gives:
\begin{align*}
	\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} B^-(1-z;b,c) \dd z 
	&=\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_0^{1-z} t^{b-1} (1-t)^{c-1} \dd t \dd z \\
	&=\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_z^1 s^{c-1} (1-s)^{b-1} \dd s \dd z
\end{align*}
Then changing the order of integration and using the substitution $u = \xi_{\alpha,\nu}/z$  and recognizing the upper incomplete $\Gamma$-function yields
\begin{align*}
	&\hspace{-30pt}\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} \int_z^1 s^{c-1} (1-s)^{b-1} \dd s \dd z\\
	&=\int_0^1 \int_0^s z^a e^{-\xi_{\alpha,\nu}/z} \dd z \, s^{c-1} (1-s)^{b-1} \dd s \\
	&=\int_0^1 \int_{\xi_{\alpha,\nu}/s}^\infty \xi_{\alpha,\nu}^{a+1} u^{-a-2} e^{-u} \dd u \, s^{c-1} (1-s)^{b-1} \dd s \\
	&= \xi_{\alpha,\nu}^{a+1} \int_0^1 \Gamma^+(-a-1,\xi_{\alpha,\nu}/s) s^{c-1} (1-s)^{b-1} \dd s.
		\numberthis \label{eq:integral_gamma_function}
\end{align*}
To compute this last integral we make use of the fact that the incomplete $\Gamma$-function has a representation in terms of Meijer's $G$-function (see Lemma~\ref{lem:gamma_meijer_G} in Appendix~\ref{sec:Meijer_G_functions})
\[
	\Gamma^+(-a-1,\xi_{\alpha,\nu}/s) = \MeijerG{2}{0}{1}{2}{1}{-a-1,0}{\frac{\xi_{\alpha,\nu}}{s}},
\] 
which holds for any $a\in \R$ and $s>0$ (that for a fixed second argument, the upper incomplete $\Gamma$-function is entire in the first argument, see \cite[pp. 899, 1032ff.]{gradshteyn2015table}). \PvdH{@Markus: What does entire mean?}  
We can now evaluate the integral in~\eqref{eq:integral_gamma_function} using several identities for Meijer's $G$-function. 
%formula 8.351.3, p. 
 %(see the wolfram functions website \url{http://functions.wolfram.com/06.06.26.0005.01}, it can also be found in the paper \url{https://arxiv.org/pdf/1507.05571}; , so we get the integral
First, inserting the expression for the incomplete Gamma-function into~\eqref{eq:integral_gamma_function} gives
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \int_0^1 s^{c-1} (1-s)^{b-1} \MeijerG{2}{0}{1}{2}{1}{-a-1,0}{\frac{\xi_{\alpha,\nu}}{s}} \dd s
\end{align*}
Next we apply the inversion identity for Meijer's $G$-function (see \cite[p. 209, 5.3.1.(9))]{erdelyi1953higher}) to get
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \int_0^1 s^{c-1} (1-s)^{b-1} \MeijerG{0}{2}{2}{1}{2+a,1}{0}{\frac{s}{\xi_{\alpha,\nu}}} \dd s
\end{align*}
This expression is actually the Euler transform of Meijer's $G$-function (see \cite[p. 214, 5.5.2.(5)]{erdelyi1953higher}) and (as the conditions $2+1<2(0+2)$ and $|\arg(\xi_{\alpha,\nu}^{-1})| < \frac{\pi}{2}$ (as $\xi_{\alpha,\nu}>0$) and $1-c-b<1-c$ (as $b>0$)  are satisfied) it equals
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \Gamma(b) \MeijerG{0}{3}{3}{2}{1-c,2+a,1}{0,1-c-b}{\xi_{\alpha,\nu}^{-1}}
\end{align*}
Using again the inversion identity for Meijer's $G$-function we now get
\begin{align*}
	\xi_{\alpha,\nu}^{a+1} \Gamma(b) \MeijerG{3}{0}{2}{3}{1,b+c}{c,-1-a,0}{\xi_{\alpha,\nu}}
\end{align*}
Finally, plugging in $a= 6\alpha-k-3$, $b=2\alpha$, $c=3-4\alpha$ we obtain
\begin{equation}\label{eq:integral_Delta_P_Beta}
	\int_0^1 z^a e^{-\xi_{\alpha,\nu}/z} B^-(1-z;b,c) \dd z =
	\xi_{\alpha,\nu}^{6a-k-2} \Gamma(2\alpha) \MeijerG{3}{0}{2}{3}{1,3-2\alpha}{3-4\alpha,-6\alpha+k+2,0}{\xi_{\alpha,\nu}}
\end{equation}
%(note that this agrees with the result in the old approach, after `bringing the power of $\xi$ into Meijer's $G$-function'.)

### 2018-11-18 concentration.tex
final short pass before sending to Tobias and Nikolaos

Next we note that the second result from Corollary~\ref{cor:approximation_triangle_integral} states that
\begin{equation}\label{eq:asymptotic_equivalence_contraint_triangle_count}
	\Exp{\widetilde{T}_{\Pcal,n}(k_n,C)} = (1+\smallO{1})\Exp{T_{\Pcal}(k_n)},
\end{equation}
so that we can conclude that the main contribution of triangles of degree $k_n$ is given by $\widetilde{T}_{\Pcal,n}(k_n, C)$. Therefore, in order to prove Proposition \ref{prop:concentration_local_clustering_P_n} it suffices to show that $\widetilde{T}_{\Pcal,n}(k_n,C)$ is sufficiently concentrated around its mean. This is done in the following proposition.

@@

To prove Proposition~\ref{prop:concentration_tilde_T_P_n} we need to understand the joint degree distribution in $G_{\Pcal,n}$ and subsequently the joint neighborhoods of two points $p, p^\prime \in \Rcal_n$. We perform the analysis in this section. The two main results, which will be the crucial technical ingredients for the proof of Proposition~\ref{prop:concentration_tilde_T_P_n}, are Lemma ?? and Lemma ??.

@@

\PvdH{The proof of this lemma follows almost the exact same steps as that of the previous one. Is there any way we could merge them?}

@@

\PvdH{Check this proof!}

>>>>

Next we note that the second result from Corollary~\ref{cor:adjusted_triangle_counting_P_n} states that
\begin{equation}\label{eq:asymptotic_equivalence_contraint_triangle_count}
	\Exp{\widetilde{T}_{\Pcal,n}(k_n,C)} = (1+\smallO{1})\Exp{T_{\Pcal}(k_n)},
\end{equation}
so that we can conclude that the main contribution of triangles of degree $k_n$ is given by $\widetilde{T}_{\Pcal,n}(k_n, C)$. Therefore, in order to prove Proposition \ref{prop:concentration_local_clustering_P_n} it suffices to show that $\widetilde{T}_{\Pcal,n}(k_n,C)$ is sufficiently concentrated around its mean. This is done in the following proposition.

@@

To prove Proposition~\ref{prop:concentration_tilde_T_P_n} we need to understand the joint degree distribution in $G_{\Pcal,n}$ and subsequently the joint neighborhoods of two points $p, p^\prime \in \Rcal_n$. We perform the analysis in this section. The two main results, which will be the crucial technical ingredients for the proof of Proposition~\ref{prop:concentration_tilde_T_P_n}, are Lemma~\ref{lem:joint_degree_factorization} and Lemma~\ref{lem:joint_degree_distribution_shift}.

@@

\PvdH{@All: The proof of this lemma follows almost the exact same steps as that of the previous one. Is there any way we could merge them?}

@@

### 2018-11-18 concentration_argument.tex
final short pass before sending to Tobias and Nikolaos

Recall that $\rho(y, k)$ is the probability density function of a Poisson random variable with expectation $\mu_{\alpha, \nu}(B_\Pcal(y)) = \xi_{\alpha,\nu}e^{\frac{y}{2}}$. 
For any sequence $\{k_n\}_{n \ge 1}$, possibly constant, define
\begin{equation}\label{eq:def_kappa_n}
	\kappa_n := \begin{cases}
		\log(n) &\mbox{if } k_n = \bigT{1},\\
		\sqrt{k_n \log(k_n)} &\mbox{else.}
	\end{cases}
\end{equation}
and define further, for any $C > 0$
\begin{equation}\label{eq:def_K_C_set}
	\Kcal_C(k_n) = \left\{p=(x,y) \in \R \times \R_+ : \frac{k_n - C \kappa_n}{\xi_{\alpha,\nu}} \vee 0 \le e^{\frac{y}{2}}
	\le \frac{k_n + C \kappa_n}{\xi_{\alpha,\nu}} \wedge e^{R_n/2} \right\},
\end{equation}
\MS{What does $\vee$ and $\wedge$ mean?}

@@

\begin{remark}[Concentration argument]\label{rmk:concentration_argument}
Lemma~\ref{lem:concentration_argument} will prove very useful in the remainder of this paper since we often have to deal with integrands of the form $g_n(y) f_{\alpha,\nu}(x,y)$ where $g_n(y) = \bigO{n^{-1} k_n^{s}}e^{\beta y}\rho(y,k_n)$ for some $s \in \R$ and $\beta < \alpha$. In this case the lemma tells us that for a suitable $C > 0$ we only need to integrate over $\Kcal_C(k_n)$. In order words, we may always assume for $g_n(y)$ (for a penalty of $\smallO{1}$) that $k_n - C \kappa_n \le \xi_{\alpha,\nu} e^{y/2} \le k_n + C \kappa_n$. We will refer to this as a \emph{concentration argument}, e.g. \emph{by a concentration argument}
\[
	\int_{\Rcal_n} g_n(y) f_{\alpha, \nu}(x,y) \dd x \dd y 
	= (1+\smallO{1}) \int_{\Kcal_C(k_n)} g_n(y) f_{\alpha, \nu}(x,y) \dd x \dd y.
\]
\PvdH{It might be a good idea to add one more example of how the concentration argument is used in the paper.}
\end{remark}

@@

\begin{proof}
Similarly to the proof of Lemma~\ref{lem:concentration_argument} we define 
\[
	\lambda_n^\pm = \ell_n \pm C \kappa_n, \quad \text{and} \quad a_n^\pm = 2 \log\left(\frac{\lambda_n^\pm}{\xi_{\alpha,\nu}}\right).
\] 
In addition we fix $\varepsilon > 0$ such that $\varepsilon < \min\{1- \frac{s}{2\alpha(2\alpha + 1)}, 1\}$. Then since $h_n(y) = \hat{\rho}_n(y,k_n) \bigO{k_n^s}$ we have
\begin{align*}
	k_n^s \int_{(1-\varepsilon)R_n}^{R_n} \hat{\rho}_n(y,k_n) e^{-\alpha y} \dd y
	&= \bigO{1} \hat{\rho}_n((1-\varepsilon)R_n,k_n) k_n^s e^{-\alpha(1-\varepsilon)R_n} \\
	&= \bigO{\hat{\rho}_n((1-\varepsilon)R_n,k_n) k_n^s n^{-2\alpha(1 - \varepsilon)}} \\
	&= \smallO{ \hat{\rho}_n((1-\varepsilon)R_n,k_n) n^{\frac{s}{2\alpha + 1} - 2\alpha(1-\varepsilon)}} = \smallO{1},
\end{align*}
where the last step follows by our choice of $\varepsilon$. Hence it is enough to prove both statement on $\Rcal_n((1-\varepsilon)R_n, R_n)$ instead of $\Rcal_n$.
\PvdH{This is where we use that $s < 2\alpha(2\alpha + 1)$. I think this can be removed by using a more careful bound but I am not sure this is needed.}

>>>>

Recall that $\rho(y, k)$ is the probability density function of a Poisson random variable with expectation $\mu_{\alpha, \nu}(B_\Pcal(y)) = \xi_{\alpha,\nu}e^{\frac{y}{2}}$. 
For any sequence $\{k_n\}_{n \ge 1}$, possibly constant, define
\begin{equation}\label{eq:def_kappa_n}
	\kappa_n := \begin{cases}
		\log(n) &\mbox{if } k_n = \bigT{1},\\
		\sqrt{k_n \log(k_n)} &\mbox{else.}
	\end{cases}
\end{equation}
and define further, for any $C > 0$
\begin{equation}\label{eq:def_K_C_set}
	\Kcal_C(k_n) = \left\{p=(x,y) \in \R \times \R_+ : \frac{k_n - C \kappa_n}{\xi_{\alpha,\nu}} \vee 0 \le e^{\frac{y}{2}}
	\le \frac{k_n + C \kappa_n}{\xi_{\alpha,\nu}} \wedge e^{R_n/2} \right\},
\end{equation}
Where we recall that $a \vee b = \max\{a,b\}$ and $a \wedge b = \min\{a,b\}$.

@@

\begin{remark}[Concentration argument]\label{rmk:concentration_argument}
Lemma~\ref{lem:concentration_argument} will prove very useful in the remainder of this paper since we often have to deal with integrands of the form $g_n(y) f_{\alpha,\nu}(x,y)$ where $g_n(y) = \bigO{n^{-1} k_n^{s}}e^{\beta y}\rho(y,k_n)$ for some $s \in \R$ and $\beta < \alpha$. In this case the lemma tells us that for a suitable $C > 0$ we only need to integrate over $\Kcal_C(k_n)$. In order words, we may always assume for $g_n(y)$ (for a penalty of $\smallO{1}$) that $k_n - C \kappa_n \le \xi_{\alpha,\nu} e^{y/2} \le k_n + C \kappa_n$. We will refer to this as a \emph{concentration argument}, e.g. \emph{by a concentration argument}
\[
	\int_{\Rcal_n} g_n(y) f_{\alpha, \nu}(x,y) \dd x \dd y 
	= (1+\smallO{1}) \int_{\Kcal_C(k_n)} g_n(y) f_{\alpha, \nu}(x,y) \dd x \dd y.
\]
\PvdH{ @All:It might be a good idea to add one more example of how the concentration argument is used in the paper. I someone finds a good example we use later on, please add it here.}
\end{remark}

@@

\begin{proof}
Similarly to the proof of Lemma~\ref{lem:concentration_argument} we define 
\[
	\lambda_n^\pm = \ell_n \pm C \kappa_n, \quad \text{and} \quad a_n^\pm = 2 \log\left(\frac{\lambda_n^\pm}{\xi_{\alpha,\nu}}\right).
\] 
In addition we fix $\varepsilon > 0$ such that $\varepsilon < \min\{1- \frac{s}{2\alpha(2\alpha + 1)}, 1\}$. Then since $h_n(y) = \hat{\rho}_n(y,k_n) \bigO{k_n^s}$ we have
\begin{align*}
	k_n^s \int_{(1-\varepsilon)R_n}^{R_n} \hat{\rho}_n(y,k_n) e^{-\alpha y} \dd y
	&= \bigO{1} \hat{\rho}_n((1-\varepsilon)R_n,k_n) k_n^s e^{-\alpha(1-\varepsilon)R_n} \\
	&= \bigO{\hat{\rho}_n((1-\varepsilon)R_n,k_n) k_n^s n^{-2\alpha(1 - \varepsilon)}} \\
	&= \smallO{ \hat{\rho}_n((1-\varepsilon)R_n,k_n) n^{\frac{s}{2\alpha + 1} - 2\alpha(1-\varepsilon)}} = \smallO{1},
\end{align*}
where the last step follows by our choice of $\varepsilon$. Hence it is enough to prove both statement on $\Rcal_n((1-\varepsilon)R_n, R_n)$ instead of $\Rcal_n$.
\PvdH{@All: This is where we use that $s < 2\alpha(2\alpha + 1)$. I think this can be removed by using a more careful bound but I am not sure this is needed.}
